Assessment of forestry road network projection methods using the “roads” package in R

### 2022-10-14 analysis/paper/paper.Rmd
paper edits

---
title: "Assessment of forestry road network projection methods using the “roads” package in R"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
---

@@

We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as our study area because forestry is the dominant resource extraction activity in this region and much of the TSA's road network is composed of forestry resource roads built to gain access to forest harvest cutblocks (Figure 1). This allows us to project road building based on forest harvest data. With approximately 55% of Revelstoke TSA's area being non-forested land (alpine, lakes, swamp, brush, rock, etc) [@brown2010], the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development. The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministryofforests]. The TSA contains wildlife species that are at risk due to their declining populations and are thought to be negatively affected by road development, such as members of the southern mountain woodland caribou population [@brown2010]. The Revelstoke TSA is an example of a landscape where road projections could be used to manage critical wildlife habitats. The British Columbia TSA boundary shapefile layer was downloaded from the British Columbia Data Catalogue [@governmentofbritishcolumbia2021].

@@

To spatially define the cost of constructing and traveling on a forestry road we developed a cost raster. We used several spatial raster layers to construct our cost surface. Following the Interior Appraisal Manual of BC, the base the cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a]. Separate costs for river and pipeline crossings were then added and major water bodies were considered barriers and assigned a very high cost. Lastly, existing roads were assigned a cost of 0 (Table 1). The cost raster produced consists of 1703936 pixels with a resolution of 100 meters by 100 meters.

Table 1. Definitions of road building costs for building the cost surface.

@@

We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e. harvest cutblocks) and the landscape connecting the two. There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees. Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape. The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the cost of building a road at any point between the two. The path to each cutblock is determined independently of the other cutblocks. The dynamic least cost paths (DLCP) method uses the same LCP algorithm but evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. The sequence is detemined by a heuristic such as the distance to the existing road. Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock. Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm [-@kruskal1956]. (See Appendix X for a more detailed explanation of these methods). In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed.

In addition to the projection methods included in the `roads` package we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>), hereafter referred to as the Hardy QGIS method. This plugin uses a similar DLCP algorithm to the `roads` package and uses the distance to the closest roads to order the cutblocks.

We used the development of roads in the Revelstoke TSA from 1990 to the present as a case study to compare the MST, DCLP and Hardy QGIS road projection methods. We also explored the effects of the cost surface resolution and different methods of selecting landing points within cutblocks by repeating the MST projections with several variations of the input parameters. We compared two resolutions for the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution. We also compared methods for locating landing points within cutblocks by using the polygon centroid and regularly spaced and random points, with point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)). The MST projection method was used for all the variations of parameters described above and an additional projection was run using the DLCP method with regularly spaced, high density landings. To allow comparisons, we also included a cutblocks only projection which did not project roads but assumed all disturbances were with in cutblocks. For the Hardy QGIS plugin we input the 100 m resolution cost raster, cutblocks, and road network data. We kept all the default settings except for the skidding distance which was set to 200m. Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^. This resulted in a total of 15 road projection variants to be compared (Table 2).

@@

In order to assess the performance of the projections, the following characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, whether or not roads are present in each pixel; 2) Road disturbance footprint, 1 km buffer around roads [@ibisch2016a]; 3) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks; 4) Road density, density of roads in meters per hectare (m/ha); 5) Distance to nearest road, distance from each raster cell to the nearest road. The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes]. Road density is an important metric that is often used when studying anthropogenic impacts on wildlife [@beazley2004]. Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package. Distance to nearest road was calculated with a fast moving window method using the getDistFromSource function of the `roads` package. All road characteristics were stored in raster format.

@@

For each projection method the aspatial projection performance varied widely across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)). The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks while the other characteristics, presence, density and distance to nearest road, performed better across the whole landscape. Across the whole landscape the the road and forestry disturbance footprints and distance to nearest roads were the closest to the observed mean, while the road presence and density characteristics were underestimated. The road density and distance to road metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (\@ref(fig:figA1)).

When comparing between projection methods we found that the density of landing points chosen within cutblocks affected the results, with the high density method underestimating the road density and road presence metrics less than the low density and centroid methods. The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results. Hardy's QGIS plugin performed well at the overall scale but tended to overestimate the amount of roads within cutblocks.

@@

```{r fig3, fig.height=5, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road characteristics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”). Values greater than 0 indicate the projection overestimated the characteristic and negative values indicate an underestimation. The cublocks only method is not defined for the road density, road presence or distance to road characteristics."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA NA klementQGIS",
                             "NA NA NA"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP regular\nhigh density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

@@

Using difference maps to highlight areas of poor projections we found several potential causes of discrepancy. In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:fig4)b). However, the cost raster was not detailed enough to reflect the costs of crossing streams or wet areas in valley bottoms and so the projections often cross the valley more than the observed roads, which typically have two parallel roads on either side of the valley. In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:fig4)b). There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:fig4)c).

```{r fig4, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of projection results for the minimum spanning tree projection with regular high density sampling to the observed road network. The primary map shows the observed distance to the nearest road minus the projected distance and the insets show close-ups of the road networks with satellite imagery to give more context. a) Projected and observed road networks follow similar patterns although projected roads cross valley bottoms more frequently. b) Projected roads do not match observed switchback roads. c) An example of a misidentified cutblock. The satellite images are from Google (2022)."}

@@

```{r fig5, fig.height=7, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method. F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "NA klementQGIS", 
                             "NA cutOnly NA"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP regular\nhigh density",
                             "Hardy QGIS", 
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

@@

As expected all the projection methods preformed worst with the road presence metric (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint performed best (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`). For the road presence metric higher sampling densities led to somewhat higher F-measure scores caused by fewer false negatives and only slightly more false positives. For the forestry and road disturbance footprint metrics, the F-measure was similar across projection methods.

@@

meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2)

matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2)

@@

```{r figA1, fig.height=10, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA klementQGIS", 
                             "NA NA NA"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP regular\nhigh density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement, resolution) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

@@

```{r ftNelson, fig.height=6, fig.width=fig_widths["two"], fig.cap="Existing roads in Fort Nelson. Cost based road projection methods are not expected to be useful in areas where roads are built for purposes other than accessing a target location at minimal cost. For example, in the Northeast corner of the Fort Nelson Timber Supply Area seismic lines show a different pattern than foresty roads."}
ftN_roads_obs <- read_sf(here("analysis/data/derived_data/combined_ft_nelson_roads.gpkg"))

>>>>

---
title: "Assessment of forestry road network projection methods using the “roads” package in R"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
---

@@

We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as our study area because forestry is the dominant resource extraction activity in this region and much of the TSA's road network is composed of forestry resource roads built to gain access to forest harvest cutblocks (Figure 1). This allows us to project road building based on forest harvest data. With approximately 55% of Revelstoke TSA's area being non-forested land (alpine, lakes, swamp, brush, rock, etc) [@brown2010], the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development. The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministryofforests]. The TSA contains wildlife species that are at risk due to their declining populations and are thought to be negatively affected by road development, such as members of the southern mountain woodland caribou population [@brown2010]. The Revelstoke TSA is an example of a landscape where road projections could be used to inform management of critical wildlife habitats.

@@

To spatially define the cost of constructing and traveling on a forestry road we developed a cost raster. We used several spatial raster layers to construct our cost surface. Following the Interior Appraisal Manual of BC, the base cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a]. Separate costs for river and pipeline crossings were then added and major water bodies were considered barriers and assigned a very high cost. Lastly, existing roads were assigned a cost of 0 (Table 1). The cost raster produced consists of 1703936 pixels with a resolution of 100 meters by 100 meters.

Table 1. Definitions of road building costs used to build the cost raster.

@@

We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e. harvest cutblocks) and the landscape connecting the two. There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees (Figure \@ref(fig:fig2)). Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape. The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the cost of building a road at any point between the two. The path to each cutblock is determined independently of the other cutblocks. The dynamic least cost paths (DLCP) method uses the same LCP algorithm but evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. The sequence is determined by a heuristic such as the distance to the existing road. Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock. Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm [-@kruskal1956]. (See Appendix X for a more detailed explanation of these methods). In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed. In addition to the projection methods included in the `roads` package, we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>), hereafter referred to as the Hardy QGIS method. This plugin uses a similar DLCP algorithm to the `roads` package and uses the distance to the closest roads to order the cutblocks. Finally, in some cases the impact of forestry on wildlife is projected without considering the road development that will accompany it, we refer to this as the cutblocks only projection method and compare it to projections that do consider roads.

@@

We used the development of roads in the Revelstoke TSA from 1990 to the present as a case study to compare the MST, DCLP and Hardy QGIS road projection methods. We also explored the effects of several variations of the input parameters by repeating the MST projections for each variant. We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution. We also compared methods for locating landing points within cutblocks by using the polygon centroid, regularly spaced pointd or random points, with point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)). The MST projection method was used for all the variations of parameters described above and an additional projection was run using the DLCP method with regularly spaced, high density landings. For the Hardy QGIS plugin we used the 100 m resolution cost raster. We kept all the default settings except for the skidding distance which was set to 200m. Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^. This resulted in a total of 15 road projection variants to be compared (Table 2).

In order to assess the performance of the projections, the following characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, whether or not roads are present in each pixel; 2) Road density, density of roads in meters per hectare (m/ha); 3) Distance to nearest road, distance from each raster cell to the nearest road. 4) Road disturbance footprint, 1 km buffer around roads [@ibisch2016a]; 5) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks; The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes]. Road density is an important metric that is often used when studying anthropogenic impacts on wildlife [@beazley2004]. Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package. Distance to nearest road was calculated with a fast moving window method using the getDistFromSource function of the `roads` package. Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable with out roads. All road characteristics were stored in raster format.

@@

For each projection method variant the aspatial performance varied across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)). The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks, while the other characteristics, presence, density and distance to nearest road, performed similarily or better across the whole landscape. Across the whole landscape the road and forestry disturbance footprints and distance to nearest roads were the closest to the observed mean, while the road presence and density characteristics were underestimated. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (\@ref(fig:figA1)).

When comparing between projection methods we found that the density of landing points chosen within cutblocks affected the results, with the high density method underestimating the road density and road presence metrics less than the low density and centroid methods. The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results. Hardy's QGIS plugin was closest to the observed mean for the road presence and density characteristics but was among the furthest from the observed mean for the distance to the nearest road and road disturbance footprint characteristic. The cutblocks only method showed that not including road development underestimates anthropogenic disturbance.

@@

```{r fig3, fig.height=3, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road characteristics using different projection method variations with a 100 m resolution cost raster across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”). Values greater than 0 indicate the projection overestimated the characteristic and negative values indicate an underestimation. The cublocks only method is not defined for the road density, road presence or distance to road characteristics."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA NA klementQGIS",
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

# save runTimes for later comparison
dlcp_time <- meanTable %>% filter(method == "dlcp", areaMean == "overall") %>% 
  pull(runTime) %>% round(2)

mst_time <- meanTable %>% filter(method == "mst", areaMean == "overall",
                                 sampleType == "regular", sampleDens == "high sample density") %>% 
  pull(runTime) %>% round(2) %>% `*`(60)

@@

Using difference maps to highlight areas of poor projections we found several potential causes of discrepancy. In most areas the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (Figure \@ref(fig:fig4)a). However, the cost raster was not detailed enough to reflect the costs of crossing streams or wet areas in valley bottoms and so the projections cross the valley more than the observed roads, which typically have two parallel roads on either side of the valley. There were also instances where cutblocks included in the data set appear to be misidentified and are not accessed by the observed road network (Figure \@ref(fig:fig4)b). In addition, the projected roads were not able to match the switchback pattern of observed roads in steep terrain (Figure \@ref(fig:fig4)c).

```{r fig4, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of projection results for the minimum spanning tree projection with regular high density sampling to the observed road network. The primary map shows the observed distance to the nearest road minus the projected distance and the insets show close-ups of the road networks with satellite imagery to give more context. a) Projected and observed road networks follow similar patterns although projected roads cross valley bottoms more frequently. b) An example of a misidentified cutblock. c) Projected roads do not match observed switchback roads.  The satellite images are from Google (2022)."}

@@

```{r fig5, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 100 m resolution cost raster. F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted. Values closer to one indicate better performance."}

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "NA klementQGIS", 
                             "NA cutOnly NA"),
                  labels = c("MST centroid", 
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS", 
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance\nfootprint"))) %>%
  group_by(metric, sampleDens, agreement) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

@@

As expected all the projection method variants preformed worst at projecting the road presence characteristic (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint had the best performance (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`). The cutblocks only method performed worst for all the road network characteristics. For the road presence metric higher sampling densities led to somewhat higher F-measure values caused by fewer false negatives and only slightly more false positives. For the forestry and road disturbance footprint metrics, the F-measure was similar across projection method variants, while sensitivity was higher at high sampling densities and precision was higher at low sampling densities. The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` for the MST method.

@@

meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2) %>% 
  filter((method != "dlcp"| is.na(method)) & (sampleType != "cutOnly" | is.na(sampleType)))

matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2) %>% 
  filter(method != "dlcp", sampleType != "cutOnly")

@@

```{r figA1, fig.height=10, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA klementQGIS", 
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "NA klementQGIS"),
                  labels = c("MST centroid", 
                             "Random low density",
                             "Regular low density",
                             "Random high density",
                             "Regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement, resolution) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

@@

```{r ftNelson, cache=TRUE, fig.height=6, fig.width=fig_widths["two"], fig.cap="Existing roads in Fort Nelson. Cost based road projection methods are not expected to be useful in areas where roads are built for purposes other than accessing a target location at minimal cost. For example, in the Northeast corner of the Fort Nelson Timber Supply Area seismic lines show a different pattern than foresty roads."}
ftN_roads_obs <- read_sf(here("analysis/data/derived_data/combined_ft_nelson_roads.gpkg"))
